\chapter*{Abstract}

Self-organizing maps, also known as Kohonen networks, are an example of artificial neural networks trained using unsupervised learning. They are not widely popular, but they are occasionally used for reducing the dimensionality of input data.This study explores the potential application of Self-Organizing Maps (SOM) networks in data classification tasks. This problem has a key role in the fields of automation and robotics, as it facilitates precise decision-making based on current information.
\\ The paper includes both a theoretical description of artificial neural networks and self-organizing maps. In addition, it presents a solution implementing SOM networks with the WTM algorithm, implemented in Python, and discusses the research conducted on three publicly available datasets: MNIST, IRIS and Cardiotocography, along with a description of the results.
\\ In the conducted research, special emphasis was placed on studying the impact of the applied neighborhood function on the learning process of the network, based on the analysis of quantization error and topographic error. The obtained results indicated the effectiveness of the SOM network in the process of input data classification, in the case of sets with a small number of dimensions. In the case of more complex sets, the results obtained were less promising, although still acceptable, indicating the possibility of conducting research in which self-organizing networks will be supplemented with other data processing techniques. 

\bigskip

\noindent\textbf{Keywords:} Self-organizing maps, unsupervised learning, clustering, classification, WTM algorithm

\bigskip

\noindent\textbf{OECD consistent field of science and technology classification:} Engineering and technology, Electrical engineering, Electronic engineering, Information engineering, Robotics and automatic control
